---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![DOI](http://joss.theoj.org/papers/10.21105/joss.01028/status.svg)](https://doi.org/10.21105/joss.01028)

# Goal

This package provides three R^2^s for statistical models with correlated errors including classes: 'lmerMod' (LMM), 'glmerMod' (GLMM), 'phylolm' (Phylogenetic GLS), and 'binaryPGLMM/phyloglm/communityPGLMM' (Phylogenetic Logistic Regression). Detailed technical descriptions can be found in [Ives 2018](https://doi.org/10.1093/sysbio/syy060).

# Installation

This package can be installed with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("arives/rr2")
```

# Package structure

This package has three main functions: `R2.resid()`, `R2.lik()`, and `R2.pred()`. You can use them individually in the form of, e.g., `R2.resid(mod, mod.r)` where `mod` is the full model and `mod.r` is the reduced model for partial R2s. If you do not include the reduced model `mod.r`, then the appropriate model with just the intercept is used to give the total R^2^. When using `R2.resid` and `R2.pred` with PGLS, you need to include the phylo object containing a phylogenetic tree, e.g., `R2.resid(mod, mod.r, phy = phy)`. 

You can calculate all three R^2^s at the same time with `R2(mod, mod.r)`. You can also specify which R^2^(s) to calculate within this function by turning off unwanted methods, e.g., `R2(mod, mod.r, resid = FALSE)` or `R2(mod, mod.r, pred = FALSE)`. 

This package also has some helper functions such as `inv.logit()`, `partialR2()`, and `partialR2adj()`.

| Models                           | Available.R2s              |
| :------------------------------- | :------------------------- |
| LM                               | partialR2, partialR2adj    |
| LM                               | R2.pred, R2.resid, R2.lik  |
| GLM                              | R2.pred, R2.resid, R2.lik  |
| LMM:  lmerMod                    | R2.pred, R2.resid, R2.lik  |
| GLMM: glmerMod                   | R2.pred, R2.resid, R2.lik  |
| PGLS: phylolm                    | R2.pred, R2.resid, R2.lik  |
| PGLMM: binaryPGLMM               | R2.pred, R2.resid, ------- |
| PGLMM: phyloglm                  | -------, --------, R2.lik  |
| PGLMM: communityPGLMM (gaussian) | R2.pred, --------, R2.lik  |
| PGLMM: communityPGLMM (binomial) | R2.pred, --------, ------- |

# Usage: calculating R^2^s for regression models

First, let's simulate data that will be used to fit various models.

```{r data-sim}
# data 
set.seed(123)
p1 <- 10; nsample <- 10; n <- p1 * nsample
d <- data.frame(x1 = rnorm(n = n), 
                x2 = rnorm(n = n), 
                u1 = rep(1:p1, each = nsample), 
                u2 = rep(1:p1, times = nsample))
d$u1 <- as.factor(d$u1); d$u2 <- as.factor(d$u2)

# LMM: y with random intercept
b1 <- 1; b2 <- -1; sd1 <- 1.5
d$y_re_intercept <- b1 * d$x1 + b2 * d$x2 + 
  rep(rnorm(n = p1, sd = sd1), each = nsample) +  # random intercept u1
  rep(rnorm(n = p1, sd = sd1), times = nsample) + # random intercept u2
  rnorm(n = n)

# LMM: y with random slope
b1 <- 0; sd1 <- 1; sd.x1 <- 2
d$y_re_slope <- b1 * d$x1 + 
  rep(rnorm(n = p1, sd = sd1), each = nsample) + # random intercept u1
  d$x1 * rep(rnorm(n = p1, sd = sd.x1), times = nsample) + # random slope u1
  rnorm(n = n)

# GLMM
b1 <- 1; sd1 <- 1.5
prob <- rr2::inv.logit(b1 * d$x1 + rep(rnorm(n = p1, sd = sd1), each = nsample)) 
# random intercept u1
d$y_binary <- rbinom(n = n, size = 1, prob = prob)

# PGLS
b1 <- 1.5; signal <- 0.7
phy <- ape::compute.brlen(ape::rtree(n = n), method = "Grafen", power = 1)
phy.x <- ape::compute.brlen(phy, method = "Grafen", power = .0001)
x_trait <- ape::rTraitCont(phy.x, model = "BM", sigma = 1)
e <- signal^0.5 * ape::rTraitCont(phy, model = "BM", sigma = 1) + (1-signal)^0.5 * rnorm(n=n)
d$x_trait <- x_trait[match(names(e), names(x_trait))]
d$y_pgls <- b1 * x_trait + e
rownames(d) <- phy$tip.label	

# Phylogenetic Logistic Regression
b1 <- 1.5; signal <- 2
e <- signal * ape::rTraitCont(phy, model = "BM", sigma = 1)
e <- e[match(phy$tip.label, names(e))]
d$y_phy_binary <- rbinom(n = n, size = 1, prob = rr2::inv.logit(b1 * d$x1 + e))

head(d)
```

Then, let's fit some models and calculate their R^2^s.

## LM

```{r lm}
library(rr2)
z.f.lm <- lm(y_re_intercept ~ x1 + x2, data = d)
z.x.lm <- lm(y_re_intercept ~ x1, data = d)
z.0.lm <- lm(y_re_intercept ~ 1, data = d)

R2(mod = z.f.lm, mod.r = z.x.lm)
partialR2(mod = z.f.lm, mod.r = z.x.lm)
partialR2adj(mod = z.f.lm, mod.r = z.x.lm)
```


## LMM

```{r lmm}
z.f.lmm <- lme4::lmer(y_re_intercept ~ x1 + x2 + (1 | u1) + (1 | u2), data = d, REML = F)
z.x.lmm <- lme4::lmer(y_re_intercept ~ x1 + (1 | u1) + (1 | u2), data = d, REML = F)
z.v.lmm <- lme4::lmer(y_re_intercept ~ 1 + (1 | u2), data = d, REML = F)
z.0.lmm <- lm(y_re_intercept ~ 1, data = d)

R2(mod = z.f.lmm, mod.r = z.x.lmm)
R2(mod = z.f.lmm, mod.r = z.v.lmm)
R2(mod = z.f.lmm, mod.r = z.0.lmm)
R2(mod = z.f.lmm) # if omit mod.r, default will be the simplest model, such as z.0.lmm here.
```

## GLMM

```{r glmm}
z.f.glmm <- lme4::glmer(y_binary ~ x1 + (1 | u1), data = d, family = "binomial")
z.x.glmm <- lme4::glmer(y_binary ~ 1 + (1 | u1), data = d, family = "binomial")
z.v.glmm <- glm(y_binary ~ x1, data = d, family = "binomial")

R2(mod = z.f.glmm, mod.r = z.x.glmm)
R2(mod = z.f.glmm, mod.r = z.v.glmm)
R2(mod = z.f.glmm)
# specify sigma2_d for R2.resid()
R2.resid(mod = z.f.glmm, mod.r = z.v.glmm, sigma2_d = "s2w")
R2.resid(mod = z.f.glmm, mod.r = z.v.glmm, sigma2_d = "NS") 
R2.resid(mod = z.f.glmm, mod.r = z.v.glmm, sigma2_d = "rNS")
```

## PGLS 

```{r pgls}
z.f.pgls <- phylolm::phylolm(y_pgls ~ x_trait, phy = phy, data = d, model = "lambda")
z.v.pgls <- lm(y_pgls ~ x_trait, data = d)

# phy is needed for phylogenetic models' R2.resid and R2.pred
R2(mod = z.f.pgls, mod.r = z.v.pgls, phy = phy)
R2(mod = z.f.pgls, phy = phy)
```

## Phylogenetic Logistic Regression

*Note*: we modified `ape::binaryPGLMM` to return necessary components for `rr2::R2()`.

```{r plog}
z.f.plog <- rr2::binaryPGLMM(y_phy_binary ~ x1, data = d, phy = phy)
z.x.plog <- rr2::binaryPGLMM(y_phy_binary ~ 1, data = d, phy = phy)
z.v.plog <- glm(y_phy_binary ~ x1, data = d, family = "binomial")

# R2.lik can't be used with binaryPGLMM because it is not a ML method
R2(mod = z.f.plog, mod.r = z.x.plog)
R2(mod = z.f.plog)

z.f.plog2 <- phylolm::phyloglm(y_phy_binary ~ x1, data = d, start.alpha = 1, phy = phy)
z.x.plog2 <- phylolm::phyloglm(y_phy_binary ~ 1, data = d, phy = phy, 
                               start.alpha = min(20, z.f.plog2$alpha))
z.v.plog2 <- glm(y_phy_binary ~ x1, data = d, family = "binomial")

# R2.resid and R2.pred do not apply for phyloglm
R2(z.f.plog2, z.x.plog2) 
# alternate
R2.lik(z.f.plog2, z.x.plog2)
```

# Contributions of predictors

We can use `rr2::R2()` to calculate partial R^2^s and compare contributions of different predictors. Here is an example using `phylolm::phyloglm()`. The same comparisons can be also applied to other types of models.

```{r comparePredictors}
z.f <- phylolm::phyloglm(y_phy_binary ~ x1 + x2, data = d, start.alpha = 1, phy = phy)
z.r1 <- phylolm::phyloglm(y_phy_binary ~ x1, data = d, start.alpha = 1, phy = phy)
z.r2 <- phylolm::phyloglm(y_phy_binary ~ x2, data = d, start.alpha = 1, phy = phy)
# total R2
R2(z.f)
# contribution of x1
R2(z.f, z.r2)
# contribution of x2
R2(z.f, z.r1)
```

It is also possible to estimate the "contribution" of correlation structrues in the model. For the above example, we can replace the phylogeny with a star phylogeny and then compare the R^2^s of the two models.

```{r}
# see the first chunk R code for the build of phy.x, a star phylogeny
z.r3 <- phylolm::phyloglm(y_phy_binary ~ x1 + x2, data = d, start.alpha = 1, phy = phy.x)
R2(z.f, z.r3)
```


# Citation

Please cite the following paper if you find this package useful:

> [Anthony Ives. 2018. R2s for Correlated Data: Phylogenetic Models, LMMs, and GLMMs. Systematic Biology, syy060.](https://doi.org/10.1093/sysbio/syy060)  
> [Anthony R. Ives and Daijiang Li (2018). rr2: An R package to calculate R^2s for regression models. The
  Journal of Open Source Software, 3(30), 1028.](https://doi.org/10.21105/joss.01028)

# Contributing

Contributions are welcome. You can either provide comments and feedback by filing an issue on Github [here](https://github.com/arives/rr2/issues) or making pull requests. It may be easier if you first open 
an issue outlining what you will do in the pull request.

Questions about the package can also be posted as issues on Github.
