---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

This package provides three R2s for statistical models with correlated errors including classes: 'lmerMod' (LMM), 'glmerMod' (GLMM), 'phylolm' (Phylogenetic GLS), and 'binaryPGLMM/phyloglm' (Phylogenetic Logistic Regression). Detailed technical descriptions can be found in [Ives 2017, preprint](http://www.biorxiv.org/content/early/2017/05/30/144170).

# Installation

This package can be installed with:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("arives/rr2")
library(rr2)
```

# Package structure

This package has three main functions: `R2.resid()`, `R2.lik()`, and `R2.pred()`. You can use them individually in the form of, e.g., `R2.resid(mod, mod.r)` where `mod` is the full model and `mod.r` is the reduce model for partial R2s. If you do not include the reduced model `mod.r`, then the appropriate model with just the intercept is used to give the total R2. When using `R2.resid` and `R2.pred` with PGLS, you need to include the phylo object containing a phylogenetic tree, e.g., `R2.resid(mod, mod.r, phy = phy)`. 

You can calculate all three R2s at the same time with `R2(mod, mod.r)`. You can also specify which R2(s) to calculate within this function by turning off unwanted methods, e.g., `R2(mod, mod.r, resid = FALSE)` or `R2(mod, mod.r, resid = FALSE)`. 

This package also has some helper functions such as `inv.logit()`, `partialR2()`, and `partialR2adj()`.

| Models                           | Available.R2s              |
| :------------------------------- | :------------------------- |
| LM                               | partialR2, partialR2adj    |
| LM                               | R2.pred, R2.resid, R2.lik  |
| GLM                              | R2.pred, R2.resid, R2.lik  |
| LMM:  lmerMod                    | R2.pred, R2.resid, R2.lik  |
| GLMM: glmerMod                   | R2.pred, R2.resid, R2.lik  |
| PGLS: phylolm                    | R2.pred, R2.resid, R2.lik  |
| PGLMM: binaryPGLMM               | R2.pred, R2.resid, ------- |
| PGLMM: phyloglm                  | -------, --------, R2.lik  |
| PGLMM: communityPGLMM (gaussian) | R2.pred, --------, R2.lik  |
| PGLMM: communityPGLMM (binomial) | R2.pred, --------, ------- |

# Usage: calculating R2s for regression models

First, let's simulate data that will be used to fit various models

```{r data-sim}
# data 
set.seed(123)
p1 <- 10; nsample <- 10; n <- p1 * nsample
d <- data.frame(x1 = rnorm(n = n), 
                x2 = rnorm(n = n), 
                u1 = rep(1:p1, each = nsample), 
                u2 = rep(1:p1, times = nsample))
d$u1 <- as.factor(d$u1); d$u2 <- as.factor(d$u2)

# LMM: y with random intercept
b1 <- 1; b2 <- -1; sd1 <- 1.5
d$y_re_intercept <- b1 * d$x1 + b2 * d$x2 + 
  rep(rnorm(n = p1, sd = sd1), each = nsample) +  # random intercept u1
  rep(rnorm(n = p1, sd = sd1), times = nsample) + # random intercept u2
  rnorm(n = n)

# LMM: y with random slope
b1 <- 0; sd1 <- 1; sd.x1 <- 2
d$y_re_slope <- b1 * d$x1 + 
  rep(rnorm(n = p1, sd = sd1), each = nsample) + # random intercept u1
  d$x1 * rep(rnorm(n = p1, sd = sd.x1), times = nsample) + # random slope u1
  rnorm(n = n)

# GLMM
b1 <- 1; sd1 <- 1.5
prob <- rr2::inv.logit(b1 * d$x1 + rep(rnorm(n = p1, sd = sd1), each = nsample)) 
# random intercept u1
d$y_binary <- rbinom(n = n, size = 1, prob = prob)

# PGLS
b1 <- 1.5; signal <- 0.7
phy <- ape::compute.brlen(ape::rtree(n = n), method = "Grafen", power = 1)
phy.x <- ape::compute.brlen(phy, method = "Grafen", power = .0001)
x_trait <- ape::rTraitCont(phy.x, model = "BM", sigma = 1)
e <- signal^0.5 * ape::rTraitCont(phy, model = "BM", sigma = 1) + (1-signal)^0.5 * rnorm(n=n)
d$x_trait <- x_trait[match(names(e), names(x_trait))]
d$y_pgls <- b1 * x_trait + e
rownames(d) <- phy$tip.label	

# Phylogenetic Logistic Regression
b1 <- 1.5; signal <- 2
e <- signal * ape::rTraitCont(phy, model = "BM", sigma = 1)
e <- e[match(phy$tip.label, names(e))]
d$y_phy_binary <- rbinom(n = n, size = 1, prob = rr2::inv.logit(b1 * d$x1 + e))

head(d)
```

Then, let's fit some models and calculate their R2s.

## LMM

```{r lmm}
library(rr2)
z.f.lmm <- lme4::lmer(y_re_intercept ~ x1 + x2 + (1 | u1) + (1 | u2), data = d, REML = F)
z.x.lmm <- lme4::lmer(y_re_intercept ~ x1 + (1 | u1) + (1 | u2), data = d, REML = F)
z.v.lmm <- lme4::lmer(y_re_intercept ~ 1 + (1 | u2), data = d, REML = F)
z.0.lmm <- lm(y_re_intercept ~ 1, data = d)

R2(mod = z.f.lmm, mod.r = z.x.lmm)
R2(mod = z.f.lmm, mod.r = z.v.lmm)
R2(mod = z.f.lmm, mod.r = z.0.lmm)
R2(mod = z.f.lmm) # if omit mod.r, default will be the simplest model, such as z.0.lmm here.
```

## GLMM

```{r glmm}
z.f.glmm <- lme4::glmer(y_binary ~ x1 + (1 | u1), data = d, family = "binomial")
z.x.glmm <- lme4::glmer(y_binary ~ 1 + (1 | u1), data = d, family = "binomial")
z.v.glmm <- glm(y_binary ~ x1, data = d, family = "binomial")

R2(mod = z.f.glmm, mod.r = z.x.glmm)
R2(mod = z.f.glmm, mod.r = z.v.glmm)
R2(mod = z.f.glmm)
```

## PGLS 

```{r pgls}
z.x.pgls <- phylolm::phylolm(y_pgls ~ 1, phy = phy, data = d, model = "lambda")
lam.x.pgls <- round(z.x.pgls$optpar, digits = 4)
z.f.pgls <- phylolm::phylolm(y_pgls ~ x_trait, phy = phy, data = d, model = "lambda", 
                             starting.value = 0.98 * lam.x.pgls + 0.01)
z.v.pgls <- lm(y_pgls ~ x_trait, data = d)

# phy is needed for phylogenetic models' R2.resid and R2.pred
R2(mod = z.f.pgls, mod.r = z.v.pgls, phy = phy)
R2(mod = z.f.pgls, phy = phy)
```

## Phylogenetic Logistic Regression

```{r plog}
z.f.plog <- rr2::binaryPGLMM(y_phy_binary ~ x1, data = d, phy = phy)
z.x.plog <- rr2::binaryPGLMM(y_phy_binary ~ 1, data = d, phy = phy)
z.v.plog <- glm(y_phy_binary ~ x1, data = d, family = "binomial")

# R2.lik can't be used with binaryPGLMM because it is not a ML method
R2(mod = z.f.plog, mod.r = z.x.plog)
R2(mod = z.f.plog)

z.f.plog2 <- phylolm::phyloglm(y_phy_binary ~ x1, data = d, start.alpha = 1, phy = phy)
z.x.plog2 <- phylolm::phyloglm(y_phy_binary ~ 1, data = d, phy = phy, 
                               start.alpha = min(20, z.f.plog2$alpha))
z.v.plog2 <- glm(y_phy_binary ~ x1, data = d, family = "binomial")

# R2.resid and R2.pred do not apply for phyloglm
R2(z.f.plog2, z.x.plog2) 
# alternate
R2.lik(z.f.plog2, z.x.plog2)
```

# Citation

Please cite the following paper if you find this package useful:

> Anthony Ives. 2017. R2s for Correlated Data: Phylogenetic Models, LMMs, and GLMMs. bioRxiv 144170. doi: <https://doi.org/10.1101/144170>

